---
title: "Bayesian Price Optimization: Estimating Price Elasticities & Choosing Profit-Maximizing
  Prices under Uncertainty"
output:
  pdf_document: default
  html_notebook: default
---

1. Setup and Libraries
We utilize a Bayesian framework to estimate demand elasticity. The primary engine is brms (interface to Stan), supported by tidyverse tools for data manipulation and visualization.

```{r}

library(readxl)
library(brms)
library(dplyr)
library(readr)
library(posterior)
library(purrr)
library(tibble)
library(ggplot2)
library(cmdstanr)

# We may need a local temp directory for our Stan program to run successfully
# There may be other workarounds that would work better without causing errors
custom_temp <- "C:/Users/Atharva/R_Temp"
if(!dir.exists(custom_temp)) dir.create(custom_temp, recursive = TRUE)

# 1. Standard R temp redirect
Sys.setenv(TMPDIR = custom_temp, TMP = custom_temp, TEMP = custom_temp)

# 2. THE FIX: Force Stan to use the ROOT of R_Temp instead of a subfolder
options(substitute_cmd = custom_temp)

# 3. Clean up the 'Truth Check'
message("Current Temp Dir: ", tempdir())

set_cmdstan_path("C:/Users/Atharva/.cmdstan/cmdstan-2.38.0")

# Enter file path as required
data <- read_excel("C:/Users/Atharva/OneDrive/Desktop/My_Work/MSc-II-Work/Microsoft Excel/Retail Price Optimization Dataset.xlsx")
dim(data)

```

2. Data Preprocessing
We transform the raw retail data into a log-log format to estimate constant elasticity. We also calculate a mean competitor price to account for cross-price effects and handle missing values to ensure model stability.

Note : Whenever competitor prices are missing or non-positive, the product’s own price is used as a fallback.

```{r}

library(dplyr)

print(class(data))
print(names(data))

print(data)

# Step 2: Assign df to a safe name (not needed)
#my_df <- as.data.frame(df)  # Ensure it's a data frame

d_filtered <- data %>%filter(!is.na(qty), !is.na(unit_price))

comp_cols <- c("comp_1", "comp_2", "comp_3")
available_comp_cols <- intersect(comp_cols, names(data))
print(paste("Available competitor columns:", paste(available_comp_cols, collapse = ", ")))

# Build data step by step
d <- data %>%
  mutate(
    qty_adj = ifelse(qty <= 0, qty + 1, qty),
    ln_qty = log(qty_adj),
    ln_price = log(unit_price)
  )

# Competitor price calculation
if (length(available_comp_cols) > 0) {
  d <- d %>%
    mutate(
      comp_mean = rowMeans(select(., all_of(available_comp_cols)), na.rm = TRUE),
      comp_mean = ifelse(is.na(comp_mean) | comp_mean <= 0, unit_price, comp_mean),
      ln_comp_mean = log(comp_mean)
    )
} else {
  d <- d %>% mutate(comp_mean = unit_price, ln_comp_mean = ln_price)
}

# Add product and month factors
d <- d %>% mutate(product_id = factor(product_id), month_ind = factor(month_year))

```

3. Global Bayesian Demand Model
We implement a Bayesian Hierarchical Linear Model. By using a log-log specification, the coefficients directly represent price elasticities.

Model Highlights :
a. Formula: $$ln\_qty \sim ln\_price + ln\_comp\_mean + (1|product\_id) + (1|month\_ind)$$

The log-log functional form in the model involves regression of natural logarithm of quantity on natural logarithm of own price and competitor price, with random intercepts for both product ID and month ID to account for unobserved heterogeneity across products (SKUs) and temporal effects.

b. Priors: We use weakly informative, industry-standard priors (e.g. Normal(-1.5, 1) for own-price elasticity) to reflect typical price-sensitive demand while preventing overfitting.

c. Estimation: Hamiltonian Monte Carlo (HMC) sampling via Stan.
adapt_delta parameter is set to 0.9 to ensure adequate sampling of posterior geometry. This specification provides robust estimation of price sensitivities while properly accounting for the nested structure of the data, yielding posterior distributions that quantify uncertainty in elasticity estimates essential for optimal pricing decisions under uncertainty.

α captures baseline demand, β measures own-price elasticity, γ measures cross-price elasticity reflecting competitor price effects, and σ represents unexplained demand variability.

Here we choose our S greater than the rows involved here to help the Monte Carlo simulations simulate the data more accurately. 

```{r}

# ==========================
# GLOBAL BAYESIAN DEMAND MODEL
# ==========================
suppressPackageStartupMessages({
  library(brms)
  library(dplyr)
  library(cmdstanr)
})


library(brms)

# Model: ln_qty ~ ln_price + ln_comp_mean + (1|product_id) + (1|month_ind)
# Note : (1|x) denotes a random intercept term 
form <- bf(ln_qty ~ ln_price + ln_comp_mean + (1 | product_id) + (1 | month_ind))

# Weakly-informative, industry-style priors
priors <- c(
  prior(normal(0, 10), class = "Intercept"),
  prior(normal(-1.5, 1), class = "b", coef = "ln_price"),      # global own-price elasticity
  prior(normal(0, 1), class = "b", coef = "ln_comp_mean"),     # global competitor elasticity
  prior(normal(0, 1), class = "sd"),                           # random effects sd
  prior(exponential(1), class = "sigma")                       # noise sd
)

options(substitute_cmd = "C:/Users/Atharva/R_Temp")

# Fit the model
suppressMessages({
  fit <- brm(
    formula = form, data = d, prior = priors, chains = 4, iter = 2500,
    warmup = 1000, cores = 4, control = list(adapt_delta = 0.95), seed = 123,
    backend = "cmdstanr", refresh = 0, silent = 2
  )
})

summary(fit)

# Save right after fitting to avoid compiling again
saveRDS(fit, "bayesian_retail_fit.rds")

```

4. Posterior Profit Optimization
This framework translates demand uncertainty into pricing decisions. We simulate the expected profit across a grid of potential prices for each SKU. By integrating over the posterior distribution, we identify the price that maximizes expected profit while accounting for estimation risk.

The core innovation lies in the profit calculation that incorporates both demand uncertainty and cost structure, where unit costs are assumed proportional to current pricing. The optimization employs Monte Carlo integration over the posterior distribution to compute expected profits, identifying the profit-maximizing price for each product. Crucially, the method quantifies uncertainty around optimal pricing decisions by computing the distribution of optimal prices across posterior draws, providing credible intervals that reflect estimation uncertainty.

The implementation scales efficiently across all products using functional programming patterns, handling edge cases where insufficient data exists for reliable estimation. The output provides a complete decision support matrix containing current pricing, recommended optimal prices with uncertainty quantification, and expected profit impacts, enabling data-driven pricing strategy with proper accounting for statistical uncertainty across the entire product catalog.

```{r}
library(posterior)
library(purrr)
library(tibble)
library(ggplot2)

posterior_draws <- as_draws_df(fit)
sigma_draws <- posterior_draws$sigma
S_total <- length(sigma_draws)

# Helper: optimal price for ONE product_id
optimize_for_sku <- function(sku_id, n_price_points = 60, n_draws = 2000) {
  d_sku <- d %>% filter(product_id == sku_id)

  if (nrow(d_sku) < 3) {
    return(tibble(
      product_id        = sku_id,
      n_obs             = nrow(d_sku),
      current_price_med = median(d_sku$unit_price),
      best_price        = NA_real_,
      best_profit       = NA_real_,
      opt_price_median  = NA_real_,
      opt_price_low95   = NA_real_,
      opt_price_high95  = NA_real_
    ))
  }

  # Assume unit cost = 60% of median price for this SKU
  cost_i <- 0.6 * median(d_sku$unit_price)

  # Price grid for this SKU
  p_min <- 0.4 * min(d_sku$unit_price)
  p_max <- 3.0 * max(d_sku$unit_price)
  price_grid <- seq(p_min, p_max, length.out = n_price_points)

  # Reference competitor level for this SKU
  ln_comp_ref <- mean(d_sku$ln_comp_mean, na.rm = TRUE)

  # Newdata for prediction (we ignore random effects with re_formula = NA)
  newdata <- tibble(
    ln_price     = log(price_grid),
    ln_comp_mean = ln_comp_ref,
    product_id   = d$product_id[1],   # dummy, ignored when re_formula = NA
    month_ind    = d$month_ind[1]     # dummy
  )

  # Posterior linear predictor (log-qty), fixed effects only
  mu_draws <- posterior_linpred(
    fit,
    newdata    = newdata,
    re_formula = NA,
    transform  = FALSE,
    ndraws     = n_draws
  ) # matrix: draws x price_grid

  S <- nrow(mu_draws)
  # Match sigma length
  if (S > S_total) {
    sig <- sample(sigma_draws, size = S, replace = TRUE)
  } else {
    sig <- sigma_draws[1:S]
  }

  # Expected qty per draw, per price (log-normal trick)
  qty_exp <- exp(mu_draws + 0.5 * (sig^2))

  # Profit per draw & price: (p - cost) * qty
  profit_draws <- sweep(qty_exp, 2, price_grid - cost_i)

  # Posterior expected profit at each price
  exp_profit <- colMeans(profit_draws)

  # Best price by expected profit
  best_idx   <- which.max(exp_profit)
  best_price <- price_grid[best_idx]
  best_profit <- exp_profit[best_idx]

  # Draw-wise optimal price (for uncertainty in optimum)
  opt_idx_per_draw   <- apply(profit_draws, 1, which.max)
  opt_price_per_draw <- price_grid[opt_idx_per_draw]

  opt_price_median <- median(opt_price_per_draw)
  opt_price_ci     <- quantile(opt_price_per_draw, probs = c(0.025, 0.975))

  tibble(
    product_id        = sku_id,
    n_obs             = nrow(d_sku),
    current_price_med = median(d_sku$unit_price),
    best_price        = best_price,
    best_profit       = best_profit,
    opt_price_median  = opt_price_median,
    opt_price_low95   = opt_price_ci[1],
    opt_price_high95  = opt_price_ci[2]
  )
}

# Run global pricing for ALL products
all_skus <- levels(d$product_id)

sku_results <- map_dfr(all_skus, optimize_for_sku)

print(sku_results)
```

5. Visualizing the Profit Surface
Finally, we visualize the "Profit vs. Price" curve for specific SKUs. The red dashed line indicates the price point where the posterior expected profit is maximized.

```{r}
example_sku <- sku_results$product_id[2] 

plot_sku <- function(sku_id) {
  d_sku <- d %>% filter(product_id == sku_id)
  cost_i <- 0.6 * median(d_sku$unit_price)

  p_min <- 0.5 * min(d_sku$unit_price)
  p_max <- 2.5 * max(d_sku$unit_price)
  price_grid <- seq(p_min, p_max, length.out = 80)
  ln_comp_ref <- mean(d_sku$ln_comp_mean, na.rm = TRUE)

  newdata <- tibble(
    ln_price     = log(price_grid),
    ln_comp_mean = ln_comp_ref,
    product_id   = d$product_id[1],
    month_ind    = d$month_ind[1]
  )

  mu_draws <- posterior_linpred(
    fit,
    newdata    = newdata,
    re_formula = NA,
    transform  = FALSE,
    ndraws     = 2000
  )
  S <- nrow(mu_draws)
  sig <- sigma_draws[1:S]

  qty_exp <- exp(mu_draws + 0.5 * (sig^2))
  profit_draws <- sweep(qty_exp, 2, price_grid - cost_i, `*`)
  exp_profit <- colMeans(profit_draws)

  best_idx   <- which.max(exp_profit)
  best_price <- price_grid[best_idx]

  df_plot <- tibble(price = price_grid, exp_profit = exp_profit)

  ggplot(df_plot, aes(x = price, y = exp_profit)) +
    geom_line(linewidth = 1, color = "steelblue") +
    geom_vline(xintercept = best_price, color = "red", linetype = "dashed") +
    labs(
      title = paste("Posterior Expected Profit vs Price for", sku_id),
      x = "Price",
      y = "Expected Profit"
    )
}

plot_sku(example_sku)

```
